---
title: "FinalProjet_STAT627"
author: Emily and Layne 
format: html
editor: visual
---

Load Libraries

```{r}
library(readxl)
library(dplyr)
library(stringr)
library(glmnet)
library(Metrics)
library(tidyr)
library(ggplot2)
```

Load in excel file

```{r}
reg_df <- read_excel("/Users/laynedibuono/Desktop/AU_Fall_2025/STAT_627/Final_Project/data/wage_df.xlsx", skip = 1)

wages_df <- reg_df |>
  rename(
    occupation     = 1, 
    total_workers  = 2,
    total_earnings = 3,
    men_workers    = 4,
    men_earnings   = 5,
    women_workers  = 6,
    women_earnings = 7
  )
```

Clean and Preprocess Data

```{r}

clean_numeric <- function(x) {
  x <- str_replace_all(x, "\\$", "")   
  x <- str_replace_all(x, ",", "")    
  x <- na_if(x, "-")                   
  as.numeric(x)
}


wages_clean <- wages_df |>
  mutate(
    total_workers  = clean_numeric(total_workers),
    men_workers    = clean_numeric(men_workers),
    women_workers  = clean_numeric(women_workers),
    total_earnings = clean_numeric(total_earnings),
    men_earnings   = clean_numeric(men_earnings),
    women_earnings = clean_numeric(women_earnings),
    pct_women      = women_workers / total_workers,
    pct_men        = men_workers / total_workers,     
    gender_gap     = men_earnings - women_earnings
  ) |>
  filter(!is.na(total_earnings))  


```

Model Matrix

```{r}
model_data <- wages_clean |>
  select(total_earnings, men_earnings, women_earnings, pct_women)

X <- model.matrix(total_earnings ~ men_earnings + women_earnings + pct_women,
                  data = model_data)[, -1]  
y <- model_data$total_earnings
```

Create OSL Baseline & RSME

```{r}
ols_model <- lm(total_earnings ~ men_earnings + women_earnings + pct_women, data = model_data)
summary(ols_model)

ols_pred <- predict(ols_model, newdata = model_data)
rmse(y, ols_pred)
```

Check for Correlation

```{r}
corr_data <- wages_clean |>
  filter(complete.cases(men_earnings, women_earnings))

cor_men_women <- cor(corr_data$men_earnings, corr_data$women_earnings)
cor_men_women
```

Plot of Correlation

```{r}
ggplot(corr_data, aes(x = men_earnings, y = women_earnings)) +
  geom_point(color = "steelblue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "purple", se = TRUE) + 
  labs(
    title = "Correlation between Men's and Women's Earnings",
    x = "Men's Median Weekly Earnings",
    y = "Women's Median Weekly Earnings"
  ) +
  theme_minimal() -> corr_plot
corr_plot
```

Filter and Check Dim and Len

```{r}
model_data <- wages_clean |>
  select(total_earnings, men_earnings, women_earnings, pct_women)

model_data <- model_data[complete.cases(model_data), ]

X <- model.matrix(total_earnings ~ men_earnings + women_earnings + pct_women,
                  data = model_data)[, -1] 
y <- model_data$total_earnings

dim(X)      
length(y)
```

Ridge Regression & RSME

```{r}
cv_ridge <- cv.glmnet(X, y, alpha = 0)
ridge_best <- glmnet(X, y, alpha = 0, lambda = cv_ridge$lambda.min)
coef(ridge_best)

```

Lasso Regression

```{r}
cv_lasso <- cv.glmnet(X, y, alpha = 1)
lasso_best <- glmnet(X, y, alpha = 1, lambda = cv_lasso$lambda.min)
coef(lasso_best)
```

Compute RSEM for OLS, Ridge, and Lasso

```{r}
ols_pred <- predict(ols_model, newdata = model_data)
rmse(y, ols_pred)

ridge_pred <- predict(ridge_best, newx = X)
rmse(y, ridge_pred[,1])

lasso_pred <- predict(lasso_best, newx = X)
rmse(y, lasso_pred[,1])
```

Table for RSME

```{r}

ols_coef <- coef(ols_model)
ridge_coef <- as.numeric(coef(ridge_best))
lasso_coef <- as.numeric(coef(lasso_best))

coef_table <- data.frame(
  Predictor = c("Intercept", "men_earnings", "women_earnings", "pct_women"),
  OLS       = ols_coef,
  Ridge     = ridge_coef,
  LASSO     = lasso_coef
)

rmse_values <- data.frame(
  Predictor = "RMSE",
  OLS       = rmse(y, ols_pred),
  Ridge     = rmse(y, ridge_pred[,1]),
  LASSO     = rmse(y, lasso_pred[,1])
)

report_table <- bind_rows(coef_table, rmse_values)
report_table
```

RMSE Plot

```{r}
coef_plot_df <- report_table |>
  filter(Predictor != "RMSE") |>
  pivot_longer(cols = OLS:LASSO, names_to = "Model", values_to = "Coefficient")

coef_plot_df
```

Plot for Comparison

```{r}
ggplot(coef_plot_df, aes(x = Predictor, y = Coefficient, fill = Model)) +
  geom_col(position = "dodge") +
  labs(
    title = "Comparison of Regression Coefficients: OLS vs Ridge vs LASSO",
    x = "Predictor",
    y = "Coefficient"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5)
  ) -> compare_plot
compare_plot
```

## Classification

Logistic regression baseline

```{r}
logreg <- glm(total_earnings ~ men_earnings + women_earnings + pct_women, data = model_data)

summary(logreg)
```

All predictors are significant based on the p-values. Model shows that the gender composition of the unit (pct_women) is a much stronger driver of total earnings differences than the raw pay levels of men or women, pointing to structural segregation or devaluation effects rather than just within-group pay gaps.

Histogram of fitted values

```{r}
ggplot(logreg$data, aes(x = logreg$fitted.values)) +
  geom_histogram( bins = 100) -> fitted_values_plot
fitted_values_plot
```

strong bimodal pattern. large peak around low values (0-200). second slightly lower peak around 400-600. Long right tail. Most predicted probabilities are low, most have fitted values under 1000, highest density below 600.

Model predicting mostly low probabilites. one group where the model is very confident the event did not occur ( near 0) and another where the predicted probability is moderate (second peak). few cases where the model predicts high probability.

```{r}
plot(logreg)
```

Random forest classifier

```{r}
library(randomForest)
```

```{r}
set.seed(123)
n <- nrow(model_data)
Z <- sample(n,n/2)
train_data <- model_data[Z,]
test_data <- model_data[-Z,]
```

```{r}
rf_model <- randomForest(total_earnings ~ men_earnings + women_earnings + pct_women, data = train_data, ntree = 500, mtry = 2, importance = TRUE)

rf_model
```

```{r}
pred_test <- predict(rf_model, newdata = test_data)
rmse_test <- sqrt(mean((pred_test - test_data$total_earnings)^2))
rsq_test  <- 1 - sum((pred_test - test_data$total_earnings)^2) / 
                 sum((test_data$total_earnings - mean(test_data$total_earnings))^2)
print(paste("Test RMSE:", round(rmse_test, 2)))
print(paste("Test R²:", round(rsq_test, 4)))
```

```{r}
library(ggplot2)

# Assuming you already have these objects from earlier:
# pred_test <- predict(rf_model, newdata = test_data)
# and test_data$total_earnings exists

plot_data <- data.frame(
  actual   = test_data$total_earnings,
  predicted = pred_test
)

ggplot(plot_data, aes(x = actual, y = predicted)) +
  geom_point(color = "#2c7bb6", alpha = 0.7, size = 1.8) +     # nice blue, slightly transparent
  geom_abline(intercept = 0, slope = 1, color = "red", linewidth = 1.2) +  # perfect line
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed", linewidth = 0.8) +  # optional: show the fitted line
  labs(
    title    = "Random Forest: Predicted vs Actual Total Earnings (Test Set)",
    subtitle = "RMSE = 68.98  |  R² = 97.74%",
    x        = "Actual Total Earnings",
    y        = "Predicted Total Earnings"
  ) +
  scale_x_continuous(labels = scales::dollar_format()) +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title    = element_text(face = "bold", size = 18),
    plot.subtitle = element_text(size = 14, color = "darkred"),
    axis.title    = element_text(face = "bold"),
    panel.grid.minor = element_blank()
  ) +
  coord_fixed(ratio = 1)   # forces square aspect ratio so the 45° line is truly 45°
```

```{r}
predictions <- predict(rf_model, newdata = test_data)
confusion_matrix <- table(predictions, test_data$total_earnings)

confusion_matrix
```

```{r}
library(e1071)
```

```{r}
model_data$total_earnings <- as.factor(model_data$total_earnings)

glimpse(model_data)
```

```{r}
set.seed(123)
n <- nrow(model_data)
Z <- sample(n,n/2)
train_data <- model_data[Z,]
test_data <- model_data[-Z,]
```

```{r}
linear_svm <- svm(total_earnings ~ men_earnings + women_earnings, data = train_data, kernel = "linear")

linear_svm
```

```{r}
radial_svm <- svm(as.factor(total_earnings) ~ men_earnings + women_earnings + pct_women, data = train_data, kernel = "radial")

radial_svm
```

Tuned model

```{r}
set.seed(1234)
S_tuned <- tune(svm,total_earnings ~ men_earnings + women_earnings + pct_women, data = model_data, ranges = list(
  cost = 10^seq(-2,3),
  kernel = c("linear", "polynomial", "sigmoid", "radial")))
summary(S_tuned)
```

Best performing model

```{r}
S_tuned$best.model
```

plot

```{r}
S_tuned$performances |> 
  ggplot(aes(x = cost, y = error, colour = kernel)) + 
  geom_line() +
  scale_x_log10() -> kernel_compare
kernel_compare
```

```{r}
ggsave("kernel_compare.png")
```

## Cross-validation

VIF

```{r}
install.packages("car")
library(car)
```

```{r}
vif(ols_model)
```

K-fold cross validation

```{r}
library(e1071)
```

```{r}
set.seed(123)
k <- 5

folds <- sample(rep(1:k, length.out = nrow(model_data)))

accuracies <- c()

for(i in 1:k) {
  train_fold <- model_data[folds != i, ]
  test_fold  <- model_data[folds == i, ]
  
  # Train SVM
  model <- svm(
    total_earnings ~ men_earnings + women_earnings + pct_women,
    data = train_fold,
    kernel = "linear",
    probability = TRUE
  )
  
  # Predictions
  pred <- predict(model, test_fold)
  
  # Accuracy for this fold
  fold_acc <- mean(pred == test_fold$total_earnings)
  accuracies[i] <- fold_acc
}

accuracies
mean(accuracies)

```

LOOCV

```{r}
library(boot)
```

```{r}
logreg2 <- glm(as.numeric(total_earnings) ~ men_earnings + women_earnings + pct_women, data = model_data)

loocv_result <- cv.glm(model_data, logreg2, K = nrow(model_data))

loocv_mse <- loocv_result$delta[1]
loocv_rmse <- sqrt(loocv_mse)

loocv_rmse
```

```{r}
model_data$total_earnings <- as.numeric(as.character(model_data$total_earnings))
model_data$men_earnings   <- as.numeric(as.character(model_data$men_earnings))
model_data$women_earnings <- as.numeric(as.character(model_data$women_earnings))
model_data$pct_women      <- as.numeric(as.character(model_data$pct_women))

mean_earnings <- mean(model_data$total_earnings, na.rm = TRUE)
sd_earnings   <- sd(model_data$total_earnings, na.rm = TRUE)

cat("Mean of total_earnings:      ", round(mean_earnings, 2), "\n")
cat("SD of total_earnings:        ", round(sd_earnings, 2), "\n")
cat("Naïve benchmark RMSE:        ", round(sd_earnings, 2), "\n")
cat("Your LOOCV RMSE:              14.50\n")
cat("Your RMSE as % of the mean:   ", round(14.50 / mean_earnings * 100, 1), "%\n")
cat("Approximate R² from LOOCV:    ", round(1 - (14.50434^2 / sd_earnings^2), 3), "\n")
```
